### 2021-07-12 response.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

@@

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

%% Consistent, changeable style for subscripts
\newcommand{\vvvar}{\mathrm{var}}
\newcommand{\wwwt}{\mathrm{wt}}

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\ry}[1]{\rx{\mathrm{#1}}} 
\newcommand{\rw}{\rx{\wwwt}}
\newcommand{\rv}{\rx{\vvvar}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ry}[1]{\Rx{\mathrm{#1}}}
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rw}{\Rx{\wwwt}}
\newcommand{\Rv}{\Rx{\vvvar}}

@@

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript. 
We have made revisions to our manuscript to address the reviewers' comments.
Overall, we have tried to provide a more thorough review of the literature and highlight the novelty of our work while acknowledging its limitations.
We have also added a new figure as suggested by Reviewer 3.
Below please find our detailed responses to reviewers.

@@

\rev{Reviewer \#1}

\revtext{This is a review of ``Roles of generation-interval distributions in shaping relative epidemic strength, speed, and control of new SARS-CoV-2 variants'' by Park and others for Royal Society Interface.  This is the latest in an ongoing series of papers (Champredon 2015, Champredon 2018, Park 2019, Dushoff 2021, ...) in which the last author has considered aspects of incubation, growth, and reproduction in SIR epidemic theory.  In this manuscript, the authors use theory to anticipate the effects of COVID-19 variants where the distribution of infectivity evolves.}

\revtext{The manuscript is excellently written. The figures are clear.  All the mathematics seems satisfactory.  Unfortunately, the manuscript does not make its case for novel results.}

Thank you for the review.
We have tried to provide a more thorough review of previous works.
As suggested by Reviewer 3, many of these concepts have been discussed within the epidemiology community with the recent emergence of new SARS-CoV-2 variants but have not been formally discussed.
We believe that the addition of a new simulations (Figure 5) and the revised text better highlight the novelty of our work.

@@

\revtext{The title is overly-broad as it suggests consideration of general forms of generation interval distributions (along the lines of Miller 2010 10.1016/j.jtbi.2009.08.007), when infect the authors study variation around the time of peak infectivity using a (very standard) $\gamma$-chain trick.  The study of generation time and $\gamma$ chains has a long history from the cited-but-unreviewed Svensson 2007, Wallinga 2007, to uncited Kenah 2008, A.  Lloyd 2001, D. Anderson 1980, Bailey 1964 (see Park 2019 and Champredon 2015).  In light of these and other works, the title and abstract fail to identify novel contributions to our understanding of epidemic theory.}

\swp{JD, we need a new title.}
Thank you for the suggestions. We have changed the title and the abstract to highlight the novelty of our work.

@@

\revtext{Although the authors are aware of the history of research on this topic, they do little to review the relevant material, leaving this manuscript feeling like a superficial re-hashing of obvious material.  The introduction doesn't provide any serious review of the theory}

Thank you for the suggestions. We now provide a review of this topic in the introduction along with a more detailed justification of the assumptions of our work:

``Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).''

We note that assuming a Gamma-distributed generation intervals is different from the standard Gamma-chain trick. We note this later in Secion 2:

``Various Gamma assumptions have been widely used in epidemic modeling, including for models of SARS-CoV-2 \citep{doi:10.1098/rsif.2020.0144}.
Although the Gamma assumption includes as a special case models that assume exponentially distributed generation intervals (when $\kappa=1$), corresponding to the SIR model \citep{anderson1991infectious}, it is different from the generation-interval distribution we get when infectious periods are Gamma distributed, another standard assumption in epidemic modeling (see \cite{roberts2007model} for discussion).''

@@

\revtext{Section 4 doesn't give references to prior work, and section 5 ignores almost all of the literature on the topic of estimating R from incidence data.}

Thank you for the suggestions. We have added references in Section 4 and a paragraph in Section 5 reviewing previous work on estimating R from incidence data.

@@

\revtext{Other comments:}

\revtext{Topic on which there is a considerable confusion in terminology in the literature.  In addition to "generation interval", terms "generation time", "transmission interval", "serial interval", and others have been used in the literature, all with related by different definitions.  This manuscripts definition "time between infection and transmission" implies a single transmission event, but Kenah pointed out that time-contraction appears when allowing multiple transmission events (see Champredon 2015).  A clearer definition up front will help readers.  Fortunately, the results don't depend on the specific definition.}

Thank you for the suggestions. We have tried to clarify what we mean by the generation interval. In particular, often-menetioned time contractions of generation intervals apply to the forward realized generation interval, which measure time between actual infection events, rather than the intrinsic generation interval, which measure the intrinsic infectiousness of infected individuals. Under strength-like interventions, the intrinsic generation interval does not vary across time, even though the realized generation interval contracts. These distinctions are clearly made in Champredon 2015, and we have tried to clarify these ideas again here (see Section 2 for a detailed discussion).

@@

\revtext{The 'speed'-'strength' distinction from the previous paper is seductive, but perhaps leaning into general quantitative illiteracy.  When one pays attention to units -- "speed" L/T,  "Growth rate" 1/T while "strength" has dimensions of force, typically, but reproduction number is a dimensionless ratio of cases per case.  Beyond time-independent final-size relations (which are themselves not robust to effects like behavior change), several researchers have argued that growth rates generally more useful than reproduction numbers (giving equivalent threshold conditions, for example) and not as prone to mis-interpretation.  Other authors in evolutionary biology and economics have argued that the reproduction number's scientific value is best salvaged using the "discounted reproduction success" (McNamara, Houston, and Collins 2001), rather than fighting of an R-r straw-man.}

\swp{I think you would be able to answer this better.}
Thank you for the suggestions. 
We agree the strict definition of ``strength'' from physics has a unit of force, and the reproduction number is unitless.
We now mention in the first paragraph that R is unitless: ``Epidemic ``strength'' is measured by the reproduction number $\RR$---a unitless quantity representing the average number of new infections caused by a typical infection.''

However, we respectfully disagree with the comment that the ``speed''-``strength'' distinction leans into general quantitative illiteracy.
As we have argued recently (https://doi.org/10.1098/rspb.2020.1556), there is a clear over-emphasis on the strength-based (R-based) perspective in the epidemiology and modeling community. 
Although the terms  ``speed'' and ``strength'' may be imprecise as noted, we believe that they can help the community move towards thinking about both r- and R-based perspectives.
We do not claim that either one is necessarily better, but rather they should be viewed as complementary approaches (since both r and R give equivalent threshold conditions).
In some cases, however, there are clear advantanges in relying on either strength- or speed-based perspectives.
For example, in the beginning of an epidemic, it is more appropriate to fix epidemic speed $r$ when evaluating uncertainties in the generation-interval distribution.
On the other hand, as we argue here, it is more appropriate to fix relative strength $\rho$, rather than relative speed $\delta$, when conditions are changing due to interventions.
We believe that our revisions, including the addition of a new figure as suggested by reviewer 3, have made these points clearer and is important to address the value of considering both perspectives to a wider audience.
We thus decided to keep the terms ``speed'' and ``strength'' in this manuscript.

@@

\revtext{Equation 1 As Kermack and McKendrick introduced it, the compartmental model is a convenient *specialization* of the integral models-- see the nice re-analysis by Breda et al., 2012.}

Thank you for pointing this out. We now cite Breda et al. We acknowledge that ``This framework ... generalizes compartmental model, such as the SEIR model''

\revtext{Equation 4 assumes the wild-type and variant co-dominate the population.  This may not be the cases when multiple variants are co-circulating, particularly during replacement events.}

Thank you for pointing this out. We now address this point:

``When more than two strains are co-circulating, the growth rate at a logit scale is expected to change over time and does not necessarily correspond to the relative speed even when all strains exhibit exponential changes (see Supplementary Text);
instead, as shown in \citep{campbell2021increased}, a multinomial logistic regression model is required to infer relative speed.''

@@

``For simplicity we neglect differences in the squared coefficient of variation and assume $\kappa_{\mathrm{wt}} = \kappa_{\mathrm{var}} = \kappa$.
While the exact shape of the distribution matters, we choose to focus on differences in the mean generation intervals to provide a qualitative understanding of how differences in generation-interval distributions between the wild type and the variant affects the relationship between relative strength and speed.
In general, we expect  differences in the squared coefficient of variation to have qualitatively similar effects as differences in the mean.
For example, a wider distribution allows for more early transmission, which in turn reduces epidemic strength given speed---this effect is qualitatively similar to that of a distribution with a shorter mean.''

\rev{Reviewer \#2}

\revtext{This is a well-written paper about the role of the generation interval distribution in the estimation of the relative strength (reproduction number) and speed
(epidemic growth rate) of a new variant of a disease. Using a renewal equation
framework, they consider inferring relative strength from relative speed and vice
versa as well as inferring relative strength from incidence data. While all of the
analysis in the paper is sound, I am concerned that the strong assumptions made
(mass-action epidemic, constant generation interval distribution, etc.) will limit
the reliability of the results in a practical application. However, they clearly
establish that assumptions about the generation interval distribution have important implications for the analysis of novel variants. While I am generally
skeptical of population-level analyses of infectious disease transmission based
on generation intervals, these are done frequently so this is a useful insight.
More detailed comments are below.}

Thank you for the review. We have tried to make the limitation and assumptions clearer throughout text.

@@

\revtext{1. (page 3, first paragraph) It is important to acknowledge that the renewal equation approach to the relationship between the generation interval
distribution, R, and the growth rate r assumes a mass-action epidemic.
When we assume a fixed generation interval distribution, we must add the
assumption that the rate of depletion of susceptibles over time is negligible and the prevalence of infection is low. One of the most important
findings of Ref 10 is that the mean generation interval contracts when the
rate of depletion of susceptibles increases. While the renewal equation
does allow for the generation interval to change over time, the analyses
in Sections 3–5 assume fixed mean generation intervals. Because variants
are likely to spread when susceptibles are being depleted by both variants
(due to cross-protection) and when the prevalence of infection is not low, these implicit assumptions place significant limitations on the practical
implications of the analysis.}

Thank you for pointing this out.
We acknowledge that we are assuming mass action.
However, a fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
This is due to an ongoing confusion between the intrinsic and realized generation interval as we explained above.
For example, as shown in Champredon et al 2015 and 2018, the standard SEIR model can be equivalently written in terms of the renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
We have added a paragraph in Section 2 that makes this distiction clearly:

``Under strength-based changes that reduce the hazard of transmission (e.g., susceptible depletion and decreased transmission rates), the instantaneous generation-interval distribution does not across change time \citep{fraser2007estimating}.
In this case, the instantaneous generation-interval distribution is also often referred to as the intrinsic generation-interval distribution \citep{champredon2015intrinsic,champredon2018two,gostic2020practical,park2020time}---
for example, the standard SEIR model can be equivalently expressed as a renewal equation with time-invariant intrinsic generation-interval distribution $g(\tau)$ as shown in \citep{champredon2018equivalence}.
The instantaneous generation-interval distribution is also different from the realized generation-interval distribution, which measures time between actual infection events (see \citep{champredon2015intrinsic}).
Previous studied have noted, in many contexts, that the realized generation intervals can contract due to susceptible depletion---a special case of strength-based changes---due to decreases in transmission over time \citep{kenah2008generation,nishiura2010time,champredon2015intrinsic}, even though the instantaneous generation-interval distribution remains unchanged in this scenario.
While the instantaneous generation-interval distribution can change under speed-based changes that disproportionately affect late transmissions (see \cite{fraser2007estimating} and Section 6--7 for detailed discussions), assuming a time-invariant instantaneous generation-interval distribution is often appropriate in the context of SARS-CoV-2, given that the dynamics of its spread have been primarily driven by strength-based changes, such as social distancing measures \citep{flaxman2020Rt} and vaccination \citep{moore2021vaccination}.
Indeed, most dynamical models of SARS-CoV-2 infections have solely relied on strength-based changes, either implicitly or explicitly assuming a time-invariant intrinsic generation-interval distribution.
Therefore, we neglect changes in the intrinsic generation-interval distribution over time for now and focus on the impact of strength-based changes on the inference of dynamics of new SARS-CoV-2 variants.
We revisit these ideas in Section 6 and compare the effects of speed-based interventions with those of strength-based interventions.''

@@

\revtext{ (page 2, lines 55–60) It would help to clarify what is meant by “strengthlike” and “speed-like” in interventions. From the examples, it appears that
speed-like interventions work by identifying potential infections quickly
(effectively ending the infectious period early while not necessarily altering
the hazard of infectious contact) and strength-like interventions work by
reducing the hazard of infectious contact (while not necessarily altering
the duration of infectiousness). A good explanation of these ideas is on
lines 228–231 on page 10.}

Thank you for the suggestion. We now define them more clearly throughout text.

@@

\revtext{In this manuscript, Park and colleagues present a thoughtful analysis of how assumptions about the generation interval distribution for wild-type and variant SARS-CoV-2 can bias estimates of the relative growth rate ("speed") and relative reproduction number ("strength") of the variant. The analysis is mathematically sound and offers a thorough investigation of an important concept that has been the subject of much discussion in the epidemiological community but, to my knowledge, hasn't been given the formal treatment it deserves until now. My comments are mostly minor and intended to make the findings more accessible and transferable to epidemiological practice.}

Thank you for the review.

\revtext{Abstract: 37: On first reading, the distinction between 'speed-like' and 'strength-like' interventions wasn't clear to me, nor why contact tracing should target r but not R, and why social distancing should target R but not r. Could you use more recognizable terms here to highlight how these interventions differ from one another? (is the key difference that contact tracing reduces the number of infectious individuals and thus the probability that a susceptible person comes into contact with an infectious person, while social distancing reduces the per-contact chance of transmission? Or something else?) This comes up again in line 57 of the Introduction.}

Thank you for pointing this out. We have tried to make these distinctions clearer throughout text.

@@

\revtext{166: I think it would be worth stating the range of Rwt and/or Rvar that yields this range for delta (I think it's something like 0.49-1.27 for the wild type, 0.78-2.04 for the variant, based on Figure 2 - but would be helpful if the reader didn't have to derive). Is the varying R what you're referring to when you speak of "underlying epidemiological conditions" in 166-167?}

Thank you for this suggestion. We clarify this when it is mentioned for the first time:

``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are negligible over a short period of time.''

@@

\revtext{Figure 4: I found this figure quite difficult to parse and not especially informative. Are the vertical bars the means of these distributions? If so, they should be labeled and referenced as such in the caption. Also in the figure 4 caption: the term ``generation intervals'' (plural) struck me as a bit strange - I think the singular would make more sense and be more consistent with what's used in the literature. Also, it would be helpful to put theta and phi in the vertical axis labels for E and F (e.g., "Strength (theta)" and "Speed (phi; 1/days)"). Unless this is observed speed and strength, meaning Rvar and rvar? In E, I'm also confused why the strength of the "constant-strength" line is increasing - shouldn't it be flat? Or what have I missed? I think some more details to refer back to the variables in the text might be helpful - for example, I think that (F) is the "Pre-intervention (black) and post-intervention speed of variant transmission (rvar) conditional on the mean generation interval of the variant (Gvar, colored)."}

Thank you for the suggestions. We have tried to make the figure caption clearer.

### 2021-07-12 response.tex

\revtext{Really what I would like to see here (Figure 4) is some simulated epidemic curves and perhaps final epidemic sizes - I think this would go much further to communicate the intuition behind how the speed-based and strength-based interventions differ than what's currently presented. One of the main conclusions of the paper seems to be that speed-based interventions could be more effective than strength-based interventions when the mean generation interval is long; presenting such simulations would be a good way to drive this point home and to give intuition for just how different speed-based vs strength-based interventions actually are.}

Thank you for the suggestions. We have added Figure 5 and some additional text to explain the differences between strength- and speed-based interventions.

@@

\revtext{264: "epidemiological dynamics" is used again here - what specifically do you mean?}

Thank you for the suggestions. We have rewritten this part:

``these biases may be assessed by considering whether estimates of relative strength appear to vary systematically with the direction of changes in the incidence of infections caused by the variant''

@@

\revtext{265: "can also lead to different conclusions about the effectiveness of interventions" - what sort of different conclusions? In which circumstances?}

We have added further explanations:

``Finally, differences in generation intervals can also lead to different conclusions about the effectiveness of interventions.
If the variant has longer generation intervals than the wild type, speed-like interventions will be relatively more effective than naive estimates would suggest. 
Conversely, strength-like intervention will be relatively more effective if the variant has shorter generation intervals.''

### 2021-07-12 newvariant.tex
done

Neglecting the (relatively slow) rate of new mutations, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.

>>>>

Neglecting the (relatively slow) rate of new mutations and assuming a mass-action epidemic, the current incidence of infection $i_x(t)$ caused by each strain $x$---either the wild type (``wt'') or the variant (``var'')---can be expressed in terms of their previous incidence $i_x(t-\tau)$ and the rate at which secondary infections are generated at time $t$ by individuals infected $\tau$ time units ago $K_x(t, \tau)$:
\begin{equation}
i_x(t) = \int_0^\infty i_x(t-\tau) K_x(t, \tau) \dtau.
\end{equation}
This framework provides a flexible way of modeling disease dynamics and generalizes compartmental model, such as the SEIR model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral,breda2012formulation, champredon2018equivalence}.
